Negative and Fractional Types

### 2012-06-27 rational_ext.tex
partial rewrite of 2.1 to emphasize the 2 levels more

\newcommand{\xcomment}[2]{\textbf{#1:~\textsl{#2}}}
\newcommand{\amr}[1]{\xcomment{Amr}{#1}}
\newcommand{\roshan}[1]{\xcomment{Roshan}{#1}}

@@

\newcommand{\arrow}[1]{\mathtt{#1}}
\newcommand{\PiL}{\ensuremath{\Pi}}

@@

\begin{verbatim}
Restructure this section to emphasize the following

@@

types b ::= 0 | 1 | b+b | b*b
values v ::= () | L v | R v | (v,v)
typing v : b

Semantics of values: 
 - a type b is interpreted as a set
 - a value v : b is an element of the set denoted by b

types b <-> b
combinators c ::= primitiveIso
               | c + c | c * c | c ; c 
typing c : b1 <-> b2

Semantics of combinators:
 - types b1 <-> b2 are interpreted as bijections 
   between the sets denoted by b1 and b2
 - a combinator c : b1 <-> b2 is interpeted as a 
   bijection between the appropriate sets

Evaluation is function application, i.e., 
  eval c v applies the function denoted by c to 
  the value denoted by v.

[If we want to use categorical language, 
we can express the above semantics using the category 
of finite sets and bijections.]

\end{verbatim}

@@

Generally, such an approach would be rather vacuous.  What makes this
pursuit different is that ... (will continue from here -- Jacques)

@@

Unlike the traditional situation with the $\lambda$-calculus, in {{Pi}},
there is a sharp distinction between data and programs.

\paragraph*{Data.} The sets of values and their types include:
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b + b | b * b 
% values, v ::= () | left v | right v | (v, v)
Types include the empty type {{0}}, the unit type {{1}}, sum types {{b1+b2}},
and products types {{b1*b2}}.  Values includes {{()}} which is the only value
of type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
and {{(v1,v2)}} which builds a value of product type. There are no values of
type {{0}}.

\paragraph*{Programs.} All the programs of {{Pi}} are special maps of type
{{v \rightarrow v}}. Specifically, they are witnesses to the following type
isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%zeroe :&  0 + b <-> b &: zeroi
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist0 :& 0 * b <-> 0 &: factor0
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
Each line of the above table introduces one or two combinators that witness
the isomorphism in the middle. Collectively the isomorphisms state that the
structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
multiplication distributes over addition. The isomorphisms are extended to
form a congruence relation by adding the following constructors that witness
equivalence and compatible closure:
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4

@@

\paragraph*{Adjoint.} 
An important property of the language is that every combinator {{c}} has an
adjoint {{c{dagger}}} that reverses the action of {{c}}. This is evident by
construction for the primitive isomorphisms. For the closure combinators, the
adjoint is homomorphic except for the case of sequencing in which the order
is reversed, i.e., {{(c1 (;) c2){dagger} = (c2{dagger}) (;) (c1{dagger}) }}.

@@

%subcode{opsem} include main
% [^ 0 ^] '= 0
% [^ 1 ^] '= 1
\end{multicols}

where {{+}} is numeric addition and {{*}} is numeric multiplication.

>>>>

\newcommand{\xcomment}[2]{\textbf{#1:~\textsl{#2}}}
\newcommand{\amr}[1]{\xcomment{Amr}{#1}}
\newcommand{\roshan}[1]{\xcomment{Roshan}{#1}}
\newcommand{\jacques}[1]{\xcomment{Jacques}{#1}}

@@

\newcommand{\arrow}[1]{\mathtt{#1}}

@@

\begin{verbatim}
(parts of rewrite that are done are removed from here)

@@

Semantics of combinators:
 - iso-types b1 <-> b2 are interpreted as bijections 
   between the sets denoted by b1 and b2
 - a combinator c : b1 <-> b2 is interpeted as a 
   bijection from set b1 to set b2.
 - there is a natural adjoint to c : b1 <-> b2 
   corresponding to the bijection c^* : b2 <-> b1

Evaluation is function application, i.e., 
  eval c v applies the function (bijection) denoted by c to 
  the value denoted by v.

[If we want to use categorical language, 
we can express the above semantics using the category 
of finite sets and bijections.]
\end{verbatim}

@@

@@

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
and these are classified by ordinary types
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b + b | b * b

But these should be regarded as largely ancilliary.  In particular we
do not treat the above values as first-class citizens.  They only occur
when we want to observe the effect of an isomorphism.

Nevertheless, these have a precise denotation:  a type {{b}} is 
interpreted as a set, and {{v : b}}  denotes that {{v}} is a member
of the set denoted by {{b}}.  As our (traditional) values are first-order,
and non-polymorphic, such a set interpretation is sound.

\paragraph*{Isomorphisms.} The terms of {{Pi}} are witnesses to type
isomorphisms.  They have (iso) types {{v \rightarrow v}}.  
Specifically, they are witnesses to the following type isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%zeroe :&  0 + b <-> b &: zeroi
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist0 :& 0 * b <-> 0 &: factor0
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
Each line of the above table introduces a pair of dual constants%
\footnote{where {{swap*}} and {{swap+}} are self-dual}
that witness the type isomorphism in the middle.  Collectively the isomorphisms
state that the structure {{(b,+,0,*,1)}} is a \emph{commutative semiring},
i.e., that each of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
multiplication distributes over addition.  These are the base 
(non-reducible) terms of the second, principal level of {{Pi}}.

It is worthwhile to remember that the above correspond to the \emph{axioms}
of commutative semirings.  However, seen as universally quantified,
orientable statements, these axioms also induce transformations of the
(traditional) values. \jacques{include code here to be explicit}

The isomorphisms are extended to form a congruence relation by adding the
following constructors that witness equivalence and compatible closure:
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4

The syntax is overloaded: we use the same symbol at the value-type level
and at the isomorphism-type level for denoting sums and products.  Hopefully
this will not cause undue confusion.

@@

\begin{definition}{(Syntax of {{Pi}})}
\label{def:Pi}
We collect our types, values, and isomorphisms, to get the full language
definition.
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= zeroe | zeroi 
%     &|& swap+ | assocl+ | assocr+ 
%     &|& identl* | identr* 
%     &|& swap* | assocl* | assocr* 
%     &|& dist0 | factor0 | dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
\end{definition}

\paragraph*{Adjoint.} 
An important property of the language is that every term {{c}} has an
adjoint {{c{dagger}}} that reverses the isomorphisms {{c}}.  This is evident by
construction for the primitive isomorphisms. For the closure combinators, the
adjoint is homomorphic except for the case of sequencing in which the order
is reversed, i.e., {{(c1 (;) c2){dagger} = (c2{dagger}) (;) (c1{dagger}) }}.

@@

%subcode{opsem} include main
% [^ 0 ^] '= 0
% [^ 1 ^] '= 1
\end{multicols}
where {{+}} is numeric addition and {{*}} is numeric multiplication.
